---
title: "Cues, Partisanship, and Willingness-to-Pay for Energy"
bibliography: references.bib
csl: american-political-science-association.csl
fig-cap-location: top
fontfamily: ebgaramond
link-bibliography: false
format:
  html:
    author:
      - name: Matthew C. Nowlin
        email: matthew.nowlin@uta.edu
        corresponding: true
        affiliations:
          - name: University of Texas at Arlington
            department: Department of Political Science
    template-partials:
      - title-metadata.html
  pdf:
    author:
      - name: "Matthew C. Nowlin^[Associate Professor, Department of Political Science, University of Texas at Arlington. Email: matthew.nowlin@uta.edu]"
    include-in-header:
      text: |
        \usepackage{setspace}
        \usepackage{dcolumn}
        \usepackage{float}
        \usepackage{amsmath}
        \usepackage{ulem}
        \addtokomafont{disposition}{\rmfamily}
        \normalem
  docx:
    author:
      - name: "Matthew C. Nowlin^[Associate Professor, Department of Political Science, University of Texas at Arlington. Email: matthew.nowlin@uta.edu]"
    reference-doc: custom-reference-doc.docx
---

```{r}
#| include: false
source("scripts/manuscript-setup.R")
```


<!-- Target journal Nature Climate Changes. guidelines:
​Brief communication:
	- Brief unreferenced abstract – 3 sentences, up to 70 words.
	- Title – up to 10 words (or 90 characters). 
	- Main text – 1,000-1,500 words, excluding abstract, references and figure legends, and contains no headings. 
	- Display items – up to 2 items, although this may be flexible at the discretion of the editor, provided the page limit is observed. 
	- Online Methods section should be included
-->

**Abstract: While not as polarized as climate change, support for energy sources, most notably fossil fuels and renewables, have been shown to have stark partisan differences.** 

The issue of climate change is one of the most polarized issues politically in the United States. Democrats and liberals are concerned and support actions to address climate change, whereas Republicans and conservatives are less concerned and less supportive. However, this polarization on climate change had not translated into polarization on renewable energy, until recently [pew]. While partisan differences exist regarding preferences for energy sources, particularly regarding fossil fuels and renewables [], overall the public is supportive of renewable energy, even as partisans have different reasons for their support. Democrats tend to support renewables, in part, because of climate change and Republicans support them for economic reasons []. Partisan and elite cues have been shown to drive opinion on climate change and energy preferences []. In light of recent rhetoric and actions on the part of the second Trump regarding fossil and renewables, it is likely that polarization among the public on energy sources will increase in response.   

To examine the potential for polarization on energy preferences, I use a survey experiment to examine respondents' prioritization of fossil fuels versus renewables as well as their willingness-to-pay (WTP) for those sources to generate electricity in response to a Trump cue or a climate change cue. Specifically, I surveyed ... US residents, and there were randomly assigned to receive one of three treatments, including a cue about Trump's critique's of renewable energy and his support for fossil fuels (n = ...), a climate change cue that stated scientists recommend moving away from fossil fuels to renewables to address climate change (n = ...), or a control condition with no cue (n = ...).   

As noted, fossil fuels and renewables tend to be polarized as these sources have been the subject of much of the political debates around energy. This trend has accelerated with the Trump, as he has long downplayed the risks of climate change, criticized renewable energy sources, and expressed for oil, gas, and coal.    

then their WTP for fossil fuels and renewable sources will be elicited using the Gabor-Granger method. 




Expectations include 
- Those in the Trump cue will favor more fossil fuels than those in the control 
- Those in the climate change cue will favor more renewables than those in the control 

- Those in the Trump cue will be WTP (more/less) for (fossil fuels/renewables) than those in the control group 
- Those in the climate change cue will be WTP (less/more) for (fossil fuels/renewables) than those in the control group 

- Conservative Republicans in the Trump cue will be WTP (more/less) for (fossil fuels/renewables) than those in the control group and other political beliefs 
- Liberal Democrats in the climate cue will be WTP (more/less) for (fossil fuels/renewables) than those in the control group and other political beliefs 

The following analysis using OLS to test the above expectations, with an emphasis on the partisan differences between liberal Democrats (libDem) and conservative Republicans interacting with the three cue conditions. The following analysis control for the respondents concern about the cost of energy and several demographic variables, including age, male, white, education, and income. 

```{r}
#| label: fig-priority-coef
#| fig-cap: "Effect of cue condition, political identity, and their interaction on the issue-priority scale (survey-weighted OLS, control condition and other/moderate identity as reference; concern.cost, age, male, white, edu, and inc included as controls but omitted from the plot)"
priority_int_tidy |>
  mutate(term_label = factor(term_label, levels = rev(unique(term_label)))) |>
  ggplot(aes(x = estimate, y = term_label, color = term_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_segment(aes(x = conf.low, xend = conf.high, yend = term_label), linewidth = 0.7) +
  geom_point(size = 2) +
  labs(x = "Coefficient (95% CI)", y = NULL, color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Conservative Republican: b = -1.22, p = 0.000
Liberal Democrat: b = 0.65, p = 0.000
Climate cue × Liberal Democrat: b = 0.35, p = 0.042
Trump cue × Conservative Republican: b = -0.41, p = 0.076

```{r}
#| label: fig-wtp-interaction
#| fig-cap: "Effect of cue condition, political identity, and their interaction on willingness-to-pay for fossil fuels and renewables (survey-weighted OLS, control condition and other/moderate identity as reference; concern.cost, age, male, white, edu, and inc included as controls but omitted from the plot)"
wtp_int_tidy |>
  mutate(term_label = factor(term_label, levels = rev(unique(term_label)))) |>
  ggplot(aes(x = estimate, y = term_label, color = term_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_segment(aes(x = conf.low, xend = conf.high, yend = term_label), linewidth = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~outcome) +
  labs(x = "Coefficient (95% CI)", y = NULL, color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Conservative Republican, WTP: Renewables: b = -6.82, p = 0.000
Liberal Democrat, WTP: Renewables: b = 5.30, p = 0.001
Trump cue, WTP: Fossil fuels: b = -2.56, p = 0.050
Climate cue × Liberal Democrat, WTP: Renewables: b = 3.66, p = 0.092
Trump cue × Conservative Republican, WTP: Fossil fuels: b = 3.23, p = 0.097


# Online Methods 
The data comes from a pool of respondents of the US public obtained from Cloud Research. Respondents were adult (18+) US residents, using Census data to match demographic characteristics including age, gender, and race/ethnicity. Weights are added to make the sample representative of the US public.

\newpage

# References