---
title: "Supplemental Materials: Cues, Partisanship, and Willingness-to-Pay for Energy"
bibliography: references.bib
csl: american-political-science-association.csl
fig-cap-location: top
fontfamily: ebgaramond
link-bibliography: false
format:
  html:
    author:
      - name: Matthew C. Nowlin
        email: matthew.nowlin@uta.edu
        corresponding: true
        affiliations:
          - name: University of Texas at Arlington
            department: Department of Political Science
    template-partials:
      - title-metadata.html
  pdf:
    author:
      - name: "Matthew C. Nowlin^[Associate Professor, Department of Political Science, University of Texas at Arlington. Email: matthew.nowlin@uta.edu]"
    include-in-header:
      text: |
        \usepackage{setspace}
        \usepackage{dcolumn}
        \usepackage{float}
        \usepackage{amsmath}
        \usepackage{ulem}
        \addtokomafont{disposition}{\rmfamily}
        \normalem
  docx:
    author:
      - name: "Matthew C. Nowlin^[Associate Professor, Department of Political Science, University of Texas at Arlington. Email: matthew.nowlin@uta.edu]"
    reference-doc: custom-reference-doc.docx
---

```{r}
#| include: false
source("scripts/supplemental-demo-setup.R")
```

# Robustness Check: Controlling for Demographics

The main text models the effect of cue condition, political identity, and their interaction on energy-source prioritization and willingness-to-pay (WTP) without any additional controls. This appendix reports two robustness checks. This section re-estimates every main-text model controlling for age, male, white, education, and income; the following section adds concern about the cost of electricity to those demographic controls. As shown below, the coefficients, their significance, and the substantive conclusions are consistent with the main text.

The results for the energy source-priority scale are shown in @tbl-priority-ols-supp-demo, parallel to the priority-scale table in the main text.

```{r}
#| label: tbl-priority-ols-supp-demo
#| tbl-cap: "Effect of cue condition, political beliefs, and their interaction on the energy source-priority scale, controlling for demographics (survey-weighted OLS; control condition and other/moderate identity as reference)"
modelsummary(
  list("Priority scale" = m_priority_int),
  coef_map = coef_map, stars = stars_map, gof_omit = gof_omit_pattern
)
```

```{r}
#| label: fig-priority-predicted-supp-demo
#| fig-cap: "Predicted energy source-priority scale (1-7) by cue condition and political identity, controlling for demographics (survey-weighted OLS predictions with 95% CIs; age, male, white, education, and income held at their survey-weighted means)"
priority_pred |>
  ggplot(aes(x = cue_condition, y = estimate, color = identity)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.4)) +
  labs(x = NULL, y = "Predicted energy-priority scale (1-7)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

As in the main text, conservative Republicans were more likely to prioritize fossil fuels (b = `r priority_conRep$b`, *p* `r priority_conRep$p`) and liberal Democrats more likely to prioritize renewables (b = `r priority_libDem$b`, *p* `r priority_libDem$p`). The interaction terms significant at *p* $\leq$ .10 in this specification are:

`r paste(priority_sig_bullets, collapse = "\n")`

The results for the WTP models are shown in @tbl-wtp-models-supp-demo, parallel to the WTP models table in the main text.

```{r}
#| label: tbl-wtp-models-supp-demo
#| tbl-cap: "Effect of cue condition, political beliefs, and their interaction on willingness-to-pay for fossil fuels (any positive WTP, and amount among those with any positive WTP) and for renewables, controlling for demographics (survey-weighted logit/OLS; control condition and other/moderate identity as reference)"
modelsummary(
  list(
    "Fossil fuels: Any WTP (logit)"    = m_wtp_fossil_participation,
    "Fossil fuels: Amount, if WTP > $0 (OLS)" = m_wtp_fossil_amount,
    "Renewables: Amount (OLS)"         = m_wtp_renewable_int
  ),
  coef_map = coef_map, stars = stars_map, gof_omit = gof_omit_pattern
)
```

```{r}
#| label: fig-wtp-fossil-predicted-supp-demo
#| fig-cap: "Predicted willingness-to-pay for fossil fuels, among respondents with any positive WTP, by cue condition and political identity, controlling for demographics (survey-weighted OLS predictions with 95% CIs; age, male, white, education, and income held at their survey-weighted means)"
fossil_amt_pred |>
  ggplot(aes(x = cue_condition, y = estimate, color = identity)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.4)) +
  labs(x = NULL, y = "Predicted WTP for fossil fuels ($)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-wtp-renewable-predicted-supp-demo
#| fig-cap: "Predicted willingness-to-pay for renewables by cue condition and political identity, controlling for demographics (survey-weighted OLS predictions with 95% CIs; age, male, white, education, and income held at their survey-weighted means)"
renewable_pred |>
  ggplot(aes(x = cue_condition, y = estimate, color = identity)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.4)) +
  labs(x = NULL, y = "Predicted WTP for renewables ($)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

For fossil fuels, conservative Republicans remained significantly more likely than the reference group to have any positive WTP (b = `r fossil_part_conRep$b`, log-odds scale, *p* `r fossil_part_conRep$p`), and liberal Democrats less likely (b = `r fossil_part_libDem$b`, *p* `r fossil_part_libDem$p`). For renewables, liberal Democrats were willing to pay \$`r wtp_libDem_renewable$b` more on average (climate condition: an additional \$`r wtp_climate_libDem_renewable$b`, *p* `r wtp_climate_libDem_renewable$p`), and conservative Republicans \$`r wtp_conRep_renewable$b_abs` less (*p* `r wtp_conRep_renewable$p`) — the same pattern, and nearly the same magnitudes, reported in the main text without this control.

**Summary.** Adding demographic controls does not change which cue x political-identity effects are significant, does not change the sign of any coefficient, and shifts coefficient magnitudes only trivially. The main text's cue and political-identity effects on energy prioritization and WTP are not an artifact of the demographic composition of respondents in each cue/identity group.

\newpage

```{r}
#| include: false
source("scripts/supplemental-setup.R")
```

# Robustness Check: Controlling for Demographics and Concern about Energy Cost

Energy affordability was a salient concern during the period of data collection: respondents' mean concern about the cost of electricity was `r concern_cost_mean` on a 0 to 10 scale. Because this concern could plausibly confound cue and political-identity effects on prioritization and WTP, this section re-estimates every main-text model controlling for demographics (age, male, white, education, and income) plus concern about energy cost. As shown below, the coefficients, their significance, and the substantive conclusions are consistent with the main text.

The results for the energy source-priority scale are shown in @tbl-priority-ols-supp, parallel to the priority-scale table in the main text.

```{r}
#| label: tbl-priority-ols-supp
#| tbl-cap: "Effect of cue condition, political beliefs, and their interaction on the energy source-priority scale, controlling for concern about energy cost (survey-weighted OLS; control condition and other/moderate identity as reference)"
modelsummary(
  list("Priority scale" = m_priority_int),
  coef_map = coef_map, stars = stars_map, gof_omit = gof_omit_pattern
)
```

```{r}
#| label: fig-priority-predicted-supp
#| fig-cap: "Predicted energy source-priority scale (1-7) by cue condition and political identity, controlling for concern about energy cost (survey-weighted OLS predictions with 95% CIs; concern about energy cost, age, male, white, education, and income held at their survey-weighted means)"
priority_pred |>
  ggplot(aes(x = cue_condition, y = estimate, color = identity)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.4)) +
  labs(x = NULL, y = "Predicted energy-priority scale (1-7)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

As in the main text, conservative Republicans were more likely to prioritize fossil fuels (b = `r priority_conRep$b`, *p* `r priority_conRep$p`) and liberal Democrats more likely to prioritize renewables (b = `r priority_libDem$b`, *p* `r priority_libDem$p`). The interaction terms significant at *p* $\leq$ .10 in this specification are:

`r paste(priority_sig_bullets, collapse = "\n")`

The results for the WTP models are shown in @tbl-wtp-models-supp, parallel to the WTP models table in the main text.

```{r}
#| label: tbl-wtp-models-supp
#| tbl-cap: "Effect of cue condition, political beliefs, and their interaction on willingness-to-pay for fossil fuels (any positive WTP, and amount among those with any positive WTP) and for renewables, controlling for concern about energy cost (survey-weighted logit/OLS; control condition and other/moderate identity as reference)"
modelsummary(
  list(
    "Fossil fuels: Any WTP (logit)"    = m_wtp_fossil_participation,
    "Fossil fuels: Amount, if WTP > $0 (OLS)" = m_wtp_fossil_amount,
    "Renewables: Amount (OLS)"         = m_wtp_renewable_int
  ),
  coef_map = coef_map, stars = stars_map, gof_omit = gof_omit_pattern
)
```

```{r}
#| label: fig-wtp-fossil-predicted-supp
#| fig-cap: "Predicted willingness-to-pay for fossil fuels, among respondents with any positive WTP, by cue condition and political identity, controlling for concern about energy cost (survey-weighted OLS predictions with 95% CIs; concern about energy cost, age, male, white, education, and income held at their survey-weighted means)"
fossil_amt_pred |>
  ggplot(aes(x = cue_condition, y = estimate, color = identity)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.4)) +
  labs(x = NULL, y = "Predicted WTP for fossil fuels ($)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-wtp-renewable-predicted-supp
#| fig-cap: "Predicted willingness-to-pay for renewables by cue condition and political identity, controlling for concern about energy cost (survey-weighted OLS predictions with 95% CIs; concern about energy cost, age, male, white, education, and income held at their survey-weighted means)"
renewable_pred |>
  ggplot(aes(x = cue_condition, y = estimate, color = identity)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.4)) +
  labs(x = NULL, y = "Predicted WTP for renewables ($)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

For fossil fuels, conservative Republicans remained significantly more likely than the reference group to have any positive WTP (b = `r fossil_part_conRep$b`, log-odds scale, *p* `r fossil_part_conRep$p`), and liberal Democrats less likely (b = `r fossil_part_libDem$b`, *p* `r fossil_part_libDem$p`). For renewables, liberal Democrats were willing to pay \$`r wtp_libDem_renewable$b` more on average (climate condition: an additional \$`r wtp_climate_libDem_renewable$b`, *p* `r wtp_climate_libDem_renewable$p`), and conservative Republicans \$`r wtp_conRep_renewable$b_abs` less (*p* `r wtp_conRep_renewable$p`) — the same pattern, and nearly the same magnitudes, reported in the main text without this control.

**Summary.** Adding concern about energy cost as a control does not change which cue x political-identity effects are significant, does not change the sign of any coefficient, and shifts coefficient magnitudes only trivially. The main text's cue and political-identity effects on energy prioritization and WTP are not an artifact of differential cost sensitivity across respondents.

\newpage

```{r}
#| include: false
source("scripts/supplemental-cost-interaction-setup.R")
```

# Robustness Check: Concern about Energy Cost Interacted with Cues and Political Identity

The prior section controls for concern about energy cost, entering it additively. This section asks whether the cost-concern effect itself varies by cue condition or political identity: every main-text model is re-estimated with concern about energy cost interacted with both cue conditions and with political identity, in addition to the demographic controls (age, male, white, education, income).

The results for the energy source-priority scale are shown in @tbl-priority-ols-cost-int.

```{r}
#| label: tbl-priority-ols-cost-int
#| tbl-cap: "Effect of cue condition, political beliefs, and concern about energy cost, including cue x cost and identity x cost interactions, on the energy source-priority scale (survey-weighted OLS; control condition and other/moderate identity as reference)"
modelsummary(
  list("Priority scale" = m_priority_int),
  coef_map = coef_map, stars = stars_map, gof_omit = gof_omit_pattern
)
```

Concern about energy cost had a significant positive effect on prioritizing fossil fuels over renewables (b = `r priority_cost$b`, *p* `r priority_cost$p`). This effect was significantly weaker for conservative Republicans (b = `r priority_conRep_cost$b`, *p* `r priority_conRep_cost$p`), consistent with cost concern already being incorporated into conservative Republicans' fossil-fuel-favoring baseline. The concern-about-cost terms significant at *p* $\leq$ .10 in this specification are:

`r priority_cost_sig_bullets`

The results for the WTP models are shown in @tbl-wtp-models-cost-int.

```{r}
#| label: tbl-wtp-models-cost-int
#| tbl-cap: "Effect of cue condition, political beliefs, and concern about energy cost, including cue x cost and identity x cost interactions, on willingness-to-pay for fossil fuels (any positive WTP, and amount among those with any positive WTP) and for renewables (survey-weighted logit/OLS; control condition and other/moderate identity as reference)"
modelsummary(
  list(
    "Fossil fuels: Any WTP (logit)"    = m_wtp_fossil_participation,
    "Fossil fuels: Amount, if WTP > $0 (OLS)" = m_wtp_fossil_amount,
    "Renewables: Amount (OLS)"         = m_wtp_renewable_int
  ),
  coef_map = coef_map, stars = stars_map, gof_omit = gof_omit_pattern
)
```

For fossil fuels, concern about energy cost lowered the odds of any positive WTP (b = `r fossil_part_cost$b`, log-odds scale, *p* `r fossil_part_cost$p`), but this penalty was significantly smaller for both liberal Democrats (b = `r fossil_part_libDem_cost$b`, *p* `r fossil_part_libDem_cost$p`) and conservative Republicans (b = `r fossil_part_conRep_cost$b`, *p* `r fossil_part_conRep_cost$p`) relative to the reference group. Among those with any positive fossil-fuel WTP, concern about energy cost lowered the amount they were willing to pay (b = `r fossil_amt_cost$b`, *p* `r fossil_amt_cost$p`), and this penalty was significantly offset under the Trump cue (b = `r fossil_amt_trump_cost$b`, *p* `r fossil_amt_trump_cost$p`). For renewables, none of the concern-about-cost terms reached significance. The concern-about-cost terms significant at *p* $\leq$ .10 in each model are:

**Fossil fuels, any positive WTP (logit):**

`r fossil_part_cost_sig_bullets`

**Fossil fuels, amount if WTP > \$0 (OLS):**

`r fossil_amt_cost_sig_bullets`

**Renewables, amount (OLS):**

`r renewable_cost_sig_bullets`

**Summary.** Concern about energy cost is not simply a nuisance control: it has a significant, partisanship-moderated relationship with fossil-fuel prioritization and WTP (weaker for both liberal Democrats and conservative Republicans than for the reference group), and its dampening effect on fossil-fuel WTP amount is significantly offset under the Trump cue. It shows no such moderated relationship with renewable-energy WTP. None of this changes the sign or significance of the main-text cue x political-identity effects (@tbl-priority-ols-supp and @tbl-wtp-models-supp), which are retained in this specification exactly as reported in the prior section.

\newpage

# References
